---
title: "NPDR identifies functional features given case/control status"
output: html_document
vignette: >
  %\VignetteIndexEntry{NPDR identifies functional features given case/control status}
  %\VignetteEngine{knitr::rmarkdown}
  \usepackage[UTF-8]{inputenc}
---

```{r message=FALSE, warnings=FALSE}
library(npdr)
library(dplyr)
library(tidyr) # pivot_longer
library(ggplot2)

theme_set(theme_minimal())
theme_update(panel.grid.minor = element_blank())
```

Let's examine the `case.control.3sets` simulated dataset provided in the NPDR package,
focusing on the train and holdout set for now.
More details on the simulation can be found [here](https://www.ncbi.nlm.nih.gov/pmc/articles/PMC4326454/).

```{r}
dat <- bind_rows(case.control.3sets[c("train", "holdout")])
functional_feats <- case.control.3sets$signal.names # known functional attributes
```

Functional features in `dat` are named `simvar*`.
Non-functional features are labeled `var*`.

## Univariate logistic regression
Perform a linear regression on all predictors, adjusted p-values, check functional hits

```{r}
# standardized beta and p-value
out_univariate <- uniReg(
  outcome = "class", dataset = dat, regression.type = "binomial"
) %>%
  data.frame()

uni_feats <- out_univariate %>%
  filter(p.adj < 0.05) %>%
  rownames()

out_univariate %>%
  slice_min(p.adj, n = 10, with_ties = FALSE)
```

## Run NPDR

If you have a reasonably-sized dataset, we recommend not using any parallelization option.

```{r}
system.time(
  out_npdr <- npdr("class", dat,
    regression.type = "binomial", attr.diff.type = "numeric-abs",
    nbd.method = "multisurf", nbd.metric = "manhattan", msurf.sd.frac = .5,
    neighbor.sampling = "none", fast.reg = F, dopar.nn = F, dopar.reg = F,
    padj.method = "bonferroni", verbose = T
  )
)

head(out_npdr)
```

Find attributes with NPDR-adjusted p-value less than 0.05.

```{r}
out_npdr %>% filter(pval.adj < .05)
```

How accurate was NPDR in detecting the underlying functional attributes?

```{r}
npdr_feats <- out_npdr %>%
  filter(pval.adj < .05) %>%
  pull(att)

cat(detectionStats(functional_feats, npdr_feats)$report)
```

```{r}
detectionStats(functional_feats, npdr_feats)$TPR
detectionStats(functional_feats, uni_feats)$TPR # nan
```

## NPDR vs. Relief
Relief scores estimated by CORElearn

```{r}
pcts <- seq(0, 1, .05)
out_corelearn <- CORElearn::attrEval(
  "class",
  data = dat,
  estimator = "ReliefFequalK",
  costMatrix = NULL,
  outputNumericSplits = FALSE,
  kNearestEqual = knnSURF(nrow(dat), .5)
) %>%
  data.frame(rrelief = .) %>%
  arrange(desc(rrelief)) %>%
  tibble::rownames_to_column("att")

detected <- function(top.pct, results.df, functional, method) {
  top.num <- floor(top.pct * nrow(results.df))
  top.vars <- results.df %>%
    data.frame() %>%
    {
      if (method == "relief") {
        top_n(., top.num, rrelief)
      } else {
        top_n(., -top.num, pval.att)
      }
    } %>%
    pull(att)
  power <- detectionStats(functional, top.vars)$TP # how many of top.pct are true
  ifelse(is.nan(power), 0, power) / length(functional)
}

compare_df <- data.frame(pcts = pcts) %>%
  mutate(
    Relief = sapply(pcts, detected,
      results.df = out_corelearn,
      functional = functional_feats,
      method = "relief"
    ),
    NPDR = sapply(pcts, detected,
      results.df = out_npdr,
      functional = functional_feats,
      method = "npdr"
    )
  ) %>%
  tidyr::pivot_longer(NPDR:Relief)

compare_df %>%
  ggplot(aes(x = pcts, y = value, color = name)) +
  geom_line(aes(linetype = name)) +
  geom_point(aes(shape = name), size = 2) +
  guides(shape = FALSE, linetype = FALSE) +
  scale_color_manual(values = c("#2887a1", "#A16928")) +
  labs(x = "Percent selected", y = "Percent correct", color = NULL) +
  ggtitle("Power to detect functional variables")
```

## auDC: area under the detection curve
```{r}
compare_df %>%
  group_by(name) %>%
  summarise(auDC = mean(value))
```

```{r echo=FALSE}
knitr::knit_exit()
```

```{r}
system.time(
  out_npdr <- npdr("class", case.control.data,
    regression.type = "binomial", attr.diff.type = "numeric-abs",
    nbd.method = "multisurf", nbd.metric = "manhattan", msurf.sd.frac = .5,
    neighbor.sampling = "none", fast.dist = T, dopar.nn = T, dopar.reg = T,
    padj.method = "bonferroni", verbose = T
  )
)
head(out_npdr)
```

```{r}
# don't expect any less than .05 for interaction simulations
# out_univariate[out_univariate[,"p.adj"]<.05,]

# attributes with npdr raw/nominal p-value less than .05
# rownames(out_npdr)[out_npdr$pval.attr<.05] # pval.attr, second column

# functional attribute detection stats
# npdr_feats <- row.names(out_npdr[out_npdr$pval.adj<.05,]) # p.adj<.05
```
